---
author:
  - name: "Nathalia Graf Grachet"
    url: https://github.com/nathaliagg/oak_metabolomics
    affiliation: University of Arizona
editor_options: 
  chunk_output_type: console
---


## Oak Metabolic Profiling


## Libraries


```{r}
library(tidyverse)
library(xcms)
doParallel::registerDoParallel(18)

## customFunnctions.R is from "https://raw.githubusercontent.com/jorainer/xcms-gnps-tools/master/customFunctions.R"
source("https://raw.githubusercontent.com/jorainer/xcms-gnps-tools/master/customFunctions.R")
```


## 1 - Import centroid data


```{r}
load("data_centroided.RData")
```


## 2 - Pre-processing 


#### 2.1 - Chromatographic peak detection


```{r}
## Set the CentWaveParam object
cwp <- CentWaveParam(
  peakwidth = c(10, 80),
  ppm = 25, # default
  snthresh = 10, # default
  noise = 1e3,
  mzdiff = -0.001
)

## peak detection
data_cent <- findChromPeaks(data_cent, param = cwp)
```


Extract information about number of peak detected per sample.


```{r}
chrom_peaks_df <- as.data.frame(chromPeaks(data_cent))

n.peaks.sample <- chrom_peaks_df %>% count(sample)

colnames(n.peaks.sample) <- c('sampleIndex', 'totalPeaksDetected')

n.peaks.sample <- merge(n.peaks.sample, pData(data_cent), by.x = "sampleIndex", by.y = 0)

n.peaks.sample[,c(1,4,2)]

write.csv(x = n.peaks.sample, file = "NumberDetectedPeaks_per_sample.csv", row.names = F)
```


#### 2.2 - Alignment


```{r}
## Group peaks (these parameters will be used in correspondence too)
mfraction = 0.3
bandwith = 30 # default
size_overlap_slices = 0.25 # default

## a - define PeakDensityParam
pdp <- PeakDensityParam(sampleGroups = data_cent$LeafType, 
                        bw = bandwith, # default 
                        minFraction = mfraction, 
                        binSize = size_overlap_slices # default
                        )
## b - Group peaks
data_cent <- groupChromPeaks(data_cent, pdp)

## Retention time correction using defaut parameters
## a - define PeakGroupsParam
pgp <- PeakGroupsParam(minFraction = mfraction)

## b - alignment
data_cent <- adjustRtime(data_cent, param = pgp)
```

```{r}
## test if object has adjusted retention time
hasAdjustedRtime(data_cent)
```

Retention time correction with either *obiwarp* or *peakGroups* is performed on all spectra including MS>1 levels if present in the [data](https://rdrr.io/bioc/xcms/man/adjustRtime-peakGroups.html)


#### 2.3 - Correspondence


```{r}
## define PeakDensityParam
pdp <- PeakDensityParam(sampleGroups= data_cent$LeafType, 
                        bw = bandwith,
                        minFraction = mfraction,
                        binSize = size_overlap_slices
                        )

## perform correspondence analysis
data_cent <- groupChromPeaks(data_cent, param=pdp)
```


#### 2.4 - Fill-in missing values


Missing values occur if no chromatographic peak was assigned to a feature either because peak detection failed, or because the corresponding ion is absent in the respective sample.


```{r}
## determine the number of missing values
number_na_i = sum(is.na(featureValues(data_cent)))
```

Fill-in missing peak data by a specified ppm (5) and expanding the mz range by mz width/4 (0.25)

```{r warning=False}
## a - define parameter
fpp <- FillChromPeaksParam(ppm = 5, expandMz = 0.25)
## b - fill in
data_cent <- fillChromPeaks(data_cent, param=fpp)
```

Determine the number of missing values after filling in:

```{r}
number_na_f = sum(is.na(featureValues(data_cent)))
## remaining number of na values
number_na_f

## determine number of filled peaks
# number_na_i - number_na_f
```


--> End of preprocessing


## 3 - Feature Correspondence


```{r}
## extract feature values after filling in
fmat_fld <- featureValues(data_cent, value="into", method="maxint")

## replace NA with zero
fmat_fld[is.na(fmat_fld)] <- 0 

## replace colnames with samplecode
fmat_fld <- as.data.frame(fmat_fld)
colnames(fmat_fld) <- as.vector(pd$SampleCode)

write.csv(file='featureCorrespondence.csv', fmat_fld)
```


## 4 - Feature Definitions


```{r}
## get feature definitions and intensities
featuresDef <- featureDefinitions(data_cent)

## merge feature definitions with correspondencce
dataTable <- merge(featuresDef, fmat_fld, by = 0, all = TRUE)
dataTable <- dataTable[, !(colnames(dataTable) %in% c("peakidx"))]
colnames(dataTable)[1] <- "Features"
write.csv(dataTable, "featureDefinitions.csv", quote = FALSE, row.names = FALSE)
```


## 5 - Saving Spectra information


```{r}
## spectra information of pre-processed data
## these data are useful for cloud plots
fdata <- fData(data_cent)
write.csv(fdata, "spectraInformation.csv")
```


## 6 - Extracting MS2


```{r}
# I modified the source function to write the feature name as the title of each MS2
source("modified_writeMgfData.R")
```

```{r warning=FALSE}
## export the individual spectra into a .mgf file
filteredMs2Spectra <- featureSpectra(data_cent, return.type = "Spectra", msLevel = 2)

filteredMs2Spectra <- clean(filteredMs2Spectra, all = TRUE)

filteredMs2Spectra <- formatSpectraForGNPS(filteredMs2Spectra) # this is one of the custom funtions

filteredMs2Spectra_consensus <- combineSpectra(filteredMs2Spectra, 
                                               fcol = "feature_id", 
                                               method = consensusSpectrum, 
                                               mzd = 0, 
                                               minProp = 0.5, 
                                               ppm = 25,
                                               intensityFun = median,
                                               mzFun = median)

mod_writeMgfDataFile(filteredMs2Spectra_consensus,
                     "ms2spectra_consensus.mgf")
```


## 7 - Save tic and bpc objects


```{r}
tic <- chromatogram(data_cent, aggregationFun = "sum")
bpc <- chromatogram(data_cent, aggregationFun = "max")
save(tic, bpc, file = 'tic_and_bpc.RData')
```


## 8 - Save pre-processed RData


```{r}
save.image(file = "xcms_pre-processed.RData")
```

